SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-02-18 derivations.tex
move custom commands to style file

\newcommand{\loadeq}[1]{%
   \ExecuteMetaData[eqn_dict.tex]{eq#1}%
}

% dcases
\usepackage{mathtools}
% for text right arrow
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{amsmath,eqparbox,xparse}

@@

% https://tex.stackexchange.com/a/34412/5764
\makeatletter
\NewDocumentCommand{\eqmathbox}{o O{c} m}{%
  \IfValueTF{#1}
    {\def\eqmathbox@##1##2{\eqmakebox[#1][#2]{$##1##2$}}}
    {\def\eqmathbox@##1##2{\eqmakebox{$##1##2$}}}
  \mathpalette\eqmathbox@{#3}
}
\makeatother

\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\minuseq}{\mathrel{{-}{=}}}
\newcommand{\ns}{N_{sp}}
\newcommand{\nr}{N_{reac}}
\newcommand{\conp}{CONP}
\newcommand{\conv}{CONV}
\newcommand{\dconp}{\ifmmode\text{for \conp,}\else for \conp,\fi}
\newcommand{\dconv}{\ifmmode\text{for \conv,}\else for \conv,\fi}
\newcommand{\Ru}{\mathcal{R}}

% simple command for equal-sized eqmathbox's on the lhs / rhs and conditional
% arguements are tag, text, alignment
\NewDocumentCommand{\lhs}{mmO{r}}{\eqmathbox[LHS_#1][#3]{#2}}
\NewDocumentCommand{\rhs}{mmO{l}}{\eqmathbox[RHS_#1][#3]{#2}}
\NewDocumentCommand{\cond}{mmO{r}}{\eqmakebox[COND_#1][#3]{#2}}
\NewDocumentCommand{\mathcond}{mmO{r}}{\eqmathbox[COND_#1][#3]{#2}}

% consistent multiline equation spacing
\NewDocumentCommand{\mathindent}{O{space}}{\hphantom{\mathrel{#1}}}

@@

%fix to dcases from here:http://tex.stackexchange.com/questions/252410/centering-in-dcases-environment/252414
\MHInternalSyntaxOn
\renewcommand{\dcases}
 {
  \MT_start_cases:nnnn
    {\quad}
    {$\m@th\displaystyle##$\hfil}
    {$\m@th\displaystyle##$\hfil}
    {\lbrace}
 }
\MHInternalSyntaxOff

@@

\begin{document}
\section{Introduction}
This document contains the derivations for \texttt{pyJac} v2.0, which generates code to evaluate the chemical source terms and analytical chemical kinetic Jacobians of constant-pressure\slash constant-volume, fixed-mass, adiabatic reactors.
Note that equations specific to the constant-pressure or constant-volume derivation will be marked with \conp~or \conv~respectively.
The derivations in this document based upon the output of the script \href{https://github.com/arghdos/SPyJac-paper/blob/master/derivations/scripts/derivations.py}{derivations.py} in the GitHub repository for this paper.
This script depends on the symbolic mathematics library SymPy~\cite{sympy}, and was tested with versions 1.0 and 1.1.1

\section{Governing equations}
\subsection{State variables}
\label{s:state}
The state vector for this derivation consists of the temperature, a thermodynamic state parameter (pressure or volume) and the number of moles of all species except the last in the model:
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns$ total species).

From the ideal gas law, the number of moles of the final species is determined as:
\begin{align}
\label{e:last_species}
n &= \frac{V P}{T \Ru} = \sum_{i=1}^{\ns}{n_i}\; , \nonumber \\
n_{\ns} &= \frac{V P}{T \Ru} - \sum_{i=1}^{\ns - 1}{n_i}
\end{align}

@@

From the ideal gas law, source terms for the volume and pressure variables may derived:
\begin{subequations}
\label{e:param_incomplete}
\begin{align}
\frac{\text{d} V }{\text{d} t } &= \frac{\Ru}{P} \left(T \frac{\text{d} n }{\text{d} t } + \frac{\text{d} T }{\text{d} t } n\right) & \text{\dconp} \\
\frac{\text{d} P }{\text{d} t } &= \frac{\Ru}{V} \left(T \frac{\text{d} n }{\text{d} t } + \frac{\text{d} T }{\text{d} t } n\right) & \text{\dconv}
\end{align}
\end{subequations}

To determine $\frac{\text{d} n }{\text{d} t }$, conservation of mass is invoked:
\begin{equation}
 \label{e:mass_conservation}
 \frac{\text{d} m }{\text{d} t } = 0 = \sum_{k=1}^{\ns}  W_{k} \frac{\text{d} n_{k} }{\text{d} t },
\end{equation}
where $W_{k}$ is the molecular weight of the $k$th species.
From~\cref{e:mass_conservation}, the source term of the last species may be written in terms of the species included in the state vector:
\begin{equation}
 \frac{\text{d} n }{\text{d} t }_{\ns} = \frac{-1}{W_{\ns}} \sum_{k=1}^{\ns - 1} W_{k} \frac{\text{d} n_{k} }{\text{d} t }
 \label{e:spec_ns}
\end{equation}
Using~\cref{e:spec_ns}, the total molar rate of change can be determined:
\begin{align}
\frac{\text{d} n }{\text{d} t } &= \sum_{k=1}^{\ns} \frac{\text{d} n_{k} }{\text{d} t }, \nonumber \\
\frac{\text{d} n }{\text{d} t } &= \sum_{k=1}^{\ns - 1} \left(1 - \frac{W_{k}}{W_{\ns}}\right) \frac{\text{d} n_{k} }{\text{d} t }
\label{e:total_molar}
\end{align}

@@

Additionally, the concentration and moles of the last species in the model (which is not a state variable) may be expanded in terms of the state variables of the system:
\begin{equation}
 \label{e:last_spec_conc}
   [C]_{\ns} = [C] - \sum_{k=1}^{\ns  - 1} [C]_{k}
\end{equation}
where $[C]$ is the total concentration:
\begin{equation}
 [C] = \frac{P}{T \Ru}
\end{equation}

Using~\cref{e:last_spec_conc,e:spec,e:spec_ns}, the temperature source terms may be expanded to contain only state variables (i.e., by removal of the last species' source term and overall production rate):
\begin{subequations}
\label{e:temperature_complete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns  - 1} \left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{p,\ns}} + \sum_{k=1}^{\ns  - 1} \left({C_{p, k}} - {C_{p,\ns}}\right) [C]_{k}} &\text{\dconp}\\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns  - 1} \left(U_{k} - \frac{W_{k} U_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{v,\ns}} + \sum_{k=1}^{\ns  - 1} \left({C_{v, k}} - {C_{v,\ns}}\right) [C]_{k}} &\text{\dconv}
\end{align}
\end{subequations}
This form of the temperature source term will be used for derivation of Jacobian terms, but~\cref{e:temperature_incomplete} will often be used as well due to its compactness.

@@

\subsection{Reaction rate expressions}
\label{s:rop_net}
The net species production rate for species $k$ is defined as:
\begin{equation}
 \label{e:spec_rop}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr} \nu_{k,i} q_{i}
\end{equation}
where $\nr$ is the number of reactions in the chemical kinetic model, $\nu_{k, i}$ the overall stoichiometric coefficient of species $k$ in reaction $i$ and $q_i$ the net rate-of-progress for reaction $i$:
\begin{align}
\nu_{k,i} &= \nu^{\prime\prime}_{k,i} - \nu^{\prime}_{k,i} \\
q_{i} &= R_{i} c_{i}
\end{align}
with $\nu^{\prime}_{k,i}$ and $\nu^{\prime\prime}_{k,i}$ the product and reactant stoichiometric cofficients (respectively) of species $k$ in reaction $i$.
The base rate-of-progress for the $i$th reversible reaction $R_{i}$ is given by:
\begin{align}
R_{i} &= {R_{f,i}} - {R_{r,i}} \label{e:ropnet}\\
{R_{f, i}} &= {k_{f, i}} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime}_{k,i}} \label{e:ropf}\\
{R_{r, i}} &= {k_{r, i}} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime\prime}_{k,i}} \label{e:ropr}
\end{align}
where ${k_{f, i}}$ and ${k_{r, i}}$ are the forward and reverse reaction coefficients (respectively) for the $i$th reaction, and the third-body\slash pressure modification $c_{i}$ is given by:
\begin{equation}
\label{e:rxn_pressure}
c_i = \begin{dcases}
  \rhs{pmod}{1}[c] &\text{for elementary reactions,} \\
  \rhs{pmod}{[X]_i}[c] &\text{for third-body enhanced reactions,} \\
  \rhs{pmod}{\frac{P_{r,i}}{1 + P_{r,i}} F_i} &\text{for unimolecular\slash~recombination fall-off reactions, and} \\
  \rhs{pmod}{\frac{1}{1 + P_{r,i}} F_i} &\text{for chemically-activated bimolecular reactions,}
  \end{dcases}
\end{equation}
where for the $i$th reaction $[X]_i$ is the third-body concentration, $P_{r,i}$ is the reduced pressure, and $F_i$ is the falloff blending factor.
These terms are defined in the following sections.

The forward reaction rate coefficient $k_{f, i}$ is given by the three-parameter Arrhenius expression:
\begin{equation}
  \label{e:arrhenius}
  {k_{f, i}} = A_{i} T^{\beta_{i}} \operatorname{exp}\left({- \frac{{E_{a}}_{i}}{T \Ru}}\right) \;,
\end{equation}
where $A_i$ is the pre-exponential factor, $\beta_i$ is the temperature exponent, and $T_{a, i}$ is the activation temperature given by $T_{a, i} = E_{a, i} / \mathcal{R}$.

@@

\subsection{Reverse rate coefficient}
By definition, the reverse rate coefficient, ${k_{r, i}}$, of irreversible reactions is zero, while reversible reactions may have non-zero ${k_{r, i}}$.
Note that in \texttt{pyJac}, reversible reactions with explicit reverse Arrhenius parameters are split into two irreversible reactions; this simplifies calculation inside the generated code, and eases comparison to Cantera~\cite{Goodwin:2015aa} which applies the same transformation.
For reversible reactions without an explicit parameterization, the reverse rate coefficient is calculated from ratio of the forward rate coefficient and the equilibrium constant:
\begin{align}
 {k_{r, i}} &= \frac{{k_{f, i}}}{{K_{c, i}}}\; , \label{e:kr}\\
 {K_{c, i}} &= {K_{p,i}}\left(\frac{P_{atm}}{T \Ru}\right)^{\sum_{k=1}^{\ns} \nu_{k,i}} \; ,\text{ and} \label{e:kc}\\
 {K_{p,i}} &= \text{exp}\left(\frac{\Delta S^{\circ}_i}{\Ru} - \frac{\Delta H^{\circ}_i}{\Ru T}\right) = \text{exp}\left(\sum_{k=1}^{\ns}\nu_{ki}\left(\frac{S^{\circ}_k}{\Ru} - \frac{H^{\circ}_k}{\Ru T}\right)\right) \label{e:kp}
\end{align}
where $P_{atm}$ is the pressure of one standard atmosphere in appropriate units.

By combining~\cref{e:kc,e:kp}, we obtain:
\begin{equation}
 \label{e:kc_in_kp}
 {K_{c, i}} = \left(\left(\frac{P_{atm}}{\Ru}\right)^{\sum_{k=1}^{\ns} \nu_{k,i}}\right) \operatorname{exp}\left({\sum_{k=1}^{\ns} \nu_{k,i} B_{k}}\right)
\end{equation}
where, $B_k$ is:
\begin{align}
 \label{e:Bk}
 B_{k} &= \frac{S^{\circ}_k}{\Ru} - \frac{H^{\circ}_k}{\Ru T} - \ln{T} \nonumber\; , \\
 B_{k} &= T \left(T \left(T \left(\frac{T a_{k,4}}{20} + \frac{a_{k,3}}{12}\right) + \frac{a_{k,2}}{6}\right) + \frac{a_{k,1}}{2}\right) \nonumber \\
       & \quad + \left(a_{k,0} - 1\right) \log{\left (T \right )} - a_{k,0} + a_{k,6} - \frac{a_{k,5}}{T}
\end{align}
from~\cref{e:hk,e:sk}.

@@

For a reaction enhanced (or diminished) by the presence of a third body, the reaction rate is modified by the third-body concentration $[X]_i$ given by
\begin{equation}
[X]_{i} = \sum_{k=1}^{\ns} \alpha_{k,i} [C]_{k} \;,
\end{equation}
where $\alpha_{k,i}$ is the third-body efficiency of species $k$ in the $i$th reaction.
For a default third-body efficiency of $\alpha_{k,i} = 1$, this may be rearranged to the compact-storage form:
\begin{equation}
 [X]_{i} = [C] + \sum_{k=1}^{\ns} \left(\alpha_{k,i} - 1\right) [C]_{k} \;,
\end{equation}
where only species with non-default third-body efficiencies must be stored in the model.
Expanding the concentration of the last species gives:
\begin{equation}
\label{e:thd_mix}
 [X]_{i}=[C] \alpha_{\ns,i} + \sum_{k=1}^{\ns  - 1} \left(- \alpha_{\ns,i} + \alpha_{k,i}\right) [C]_{k}\;.
\end{equation}
This form will be denoted as the mixture-based (or \textbf{mix} for short) third-body efficiency.

If all species in the mixture contribute equally as third bodies, then $\alpha_{k,i} = 1$ for all species:
\begin{equation}
\label{e:thd_unity}
 [X]_{i} = [C] = \frac{P}{\Ru T} \;.
\end{equation}
This form will be called the \textbf{unity}-based third-body efficiency.

In addition, a single species may act as the third body in which case
\begin{equation}
 [X]_{i} = [C_m]
\end{equation}
where the $m$th species is the third body.
For evaluation of Jacobian entries, the expanded form:
\begin{equation}
\label{e:thd_spec}
 [X]_{i}=\left([C] - \sum_{k=1}^{\ns  - 1} [C]_{k}\right) \delta_{\ns m} + \left(- \delta_{\ns m} + 1\right) [C]_{m}
\end{equation}
will be used, where the Kronecker delta $\delta_{\ns m}$ is unity if and only if the third body species $m$ is the last species in the model.
\Cref{e:thd_spec} will be referred to as the \textbf{species}-based third-body efficiency.

@@

\subsection{Temperature source term derivatives}
\label{s:energy_derivs}
\subsubsection{Temperature derivative}
\label{s:denergy_dt}
First, we will derive the term:
\begin{align}
 \mathcal{J}_{1,1} &= \frac{\partial}{\partial T} \frac{\text{d} T}{\text{d} t} \nonumber \\
 \intertext{or more simply:}
 \label{e:dTdotdT_initial}
 \mathcal{J}_{1,1} &= \frac{\partial\dot{T}}{\partial{T}}
\end{align}
To do so, we will more closely follow the \conp~energy equation derivative with respect to temperature as the process is largely identical between \conp~and \conv.

The derivative of a species concentration with respect to temperature for both \conp~and \conv~is:
\begin{align}
 \frac{\partial [C]_{k}}{\partial T} =
 \begin{dcases}
  \frac{\partial}{\partial T} \left(\frac{n_k}{V}\right) = 0 & \left\{k \ne \ns\right\} \\
  \frac{\partial}{\partial T} \left([C] - \sum_{k=1}^{\ns - 1}{\frac{n_k}{V}}\right) = -\frac{[C]}{T} & \left\{k = \ns\right\}
 \end{dcases}
\end{align}
as the pressure and volume are either constants (for \conp~and \conv~respectively) or a state variable and a function of time only.

@@

To make the result more managable, the denominator of~\cref{e:temperature_complete} is first collapsed back into the form of~\cref{e:temperature_incomplete}, yielding for~\conp:

@@

 \frac{\partial\dot{T}}{\partial{T}} &= - \frac{\left( [C] \left( \frac{{C_p}_{\ns}}{T} - \frac{\text{d} {C_p} }{\text{d} T }_{\ns} \right) - \sum_{k=1}^{-1 + \ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns} + \frac{\text{d} C_{p, k} }{\text{d} T }\right) [C]_{k}\right)}{\left(\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}\right)^{2}} \times \nonumber \\
				     & \mathindent[largerindent] \sum_{k=1}^{-1 + \ns} \left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k} - \nonumber \\
				     & \mathindent \frac{\sum_{k=1}^{-1 + \ns} \left(\left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega}_{k} }{\partial T } + \left(\frac{\text{d} H_{k} }{\text{d} T } - \frac{W_{k}}{W_{\ns}} \frac{\text{d} H }{\text{d} T }_{\ns}\right) \dot{\omega}_{k}\right)}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}}

@@

 \frac{\partial\dot{T}}{\partial{T}} &= \frac{1}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \times \nonumber \\
				     & \left( [C] \left( \frac{{C_p}_{\ns}}{T} - \frac{\text{d} {C_p} }{\text{d} T }_{\ns} \right) - \sum_{k=1}^{-1 + \ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns} + \frac{\text{d} {C_p}_{k} }{\text{d} T }\right) [C]_{k}\right) - \nonumber \\
				     & \sum_{k=1}^{-1 + \ns} \left(\left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega}_{k} }{\partial T } + \left(\frac{\text{d} H_{k} }{\text{d} T } - \frac{W_{k}}{W_{\ns}} \frac{\text{d} H }{\text{d} T }_{\ns}\right) \dot{\omega}_{k}\right) \Biggr]

@@

 [C] = \sum_{k=1}^{\ns} [C]_{k}

>>>>

% dcases
\usepackage{mathtools}
% for text right arrow
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{amsmath,eqparbox,xparse}
\usepackage{macros}

@@

@@

@@

\begin{document}
\section{Introduction}
This document contains the derivations for \texttt{pyJac} v2.0, which generates code to evaluate the chemical source terms and analytical chemical kinetic Jacobians of constant-pressure\slash constant-volume, fixed-mass, adiabatic reactors.
Note that equations specific to the constant-pressure or constant-volume derivation will be marked with \conp/ or \conv/ respectively.
The derivations in this document based upon the output of the script \href{https://github.com/arghdos/SPyJac-paper/blob/master/derivations/scripts/derivations.py}{derivations.py} in the GitHub repository for this paper.
This script depends on the symbolic mathematics library SymPy~\cite{sympy}, and was tested with versions 1.0 and 1.1.1

\section{Governing equations}
\subsection{State variables}
\label{s:state}
The state vector for this derivation consists of the temperature, a thermodynamic state parameter (pressure or volume) and the number of moles of all species except the last in the model:
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns/$ total species).

From the ideal gas law, the number of moles of the final species is determined as:
\begin{align}
\label{e:last_species}
n &= \frac{V P}{T \Ru/} = \sum_{i=1}^{\ns/}{n_i}\; , \nonumber \\
n_{\ns/} &= \frac{V P}{T \Ru/} - \sum_{i=1}^{\ns/ - 1}{n_i}
\end{align}

@@

From the ideal gas law, source terms for the volume and pressure variables may derived:
\begin{subequations}
\label{e:param_incomplete}
\begin{align}
\frac{\text{d} V }{\text{d} t } &= \frac{\Ru/}{P} \left(T \frac{\text{d} n }{\text{d} t } + \frac{\text{d} T }{\text{d} t } n\right) & \text{\dconp/} \\
\frac{\text{d} P }{\text{d} t } &= \frac{\Ru/}{V} \left(T \frac{\text{d} n }{\text{d} t } + \frac{\text{d} T }{\text{d} t } n\right) & \text{\dconv/}
\end{align}
\end{subequations}

To determine $\frac{\text{d} n }{\text{d} t }$, conservation of mass is invoked:
\begin{equation}
 \label{e:mass_conservation}
 \frac{\text{d} m }{\text{d} t } = 0 = \sum_{k=1}^{\ns/}  W_{k} \frac{\text{d} n_{k} }{\text{d} t },
\end{equation}
where $W_{k}$ is the molecular weight of the $k$th species.
From~\cref{e:mass_conservation}, the source term of the last species may be written in terms of the species included in the state vector:
\begin{equation}
 \frac{\text{d} n }{\text{d} t }_{\ns/} = \frac{-1}{W_{\ns/}} \sum_{k=1}^{\ns/ - 1} W_{k} \frac{\text{d} n_{k} }{\text{d} t }
 \label{e:spec_ns}
\end{equation}
Using~\cref{e:spec_ns}, the total molar rate of change can be determined:
\begin{align}
\frac{\text{d} n }{\text{d} t } &= \sum_{k=1}^{\ns/} \frac{\text{d} n_{k} }{\text{d} t }, \nonumber \\
\frac{\text{d} n }{\text{d} t } &= \sum_{k=1}^{\ns/ - 1} \left(1 - \frac{W_{k}}{W_{\ns/}}\right) \frac{\text{d} n_{k} }{\text{d} t }
\label{e:total_molar}
\end{align}

@@

Additionally, the concentration and moles of the last species in the model (which is not a state variable) may be expanded in terms of the state variables of the system:
\begin{equation}
 \label{e:last_spec_conc}
   [C]_{\ns/} = [C] - \sum_{k=1}^{\ns/  - 1} [C]_{k}
\end{equation}
where $[C]$ is the total concentration:
\begin{equation}
 [C] = \frac{P}{T \Ru/}
\end{equation}

Using~\cref{e:last_spec_conc,e:spec,e:spec_ns}, the temperature source terms may be expanded to contain only state variables (i.e., by removal of the last species' source term and overall production rate):
\begin{subequations}
\label{e:temperature_complete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns/  - 1} \left(H_{k} - \frac{W_{k} H_{\ns/}}{W_{\ns/}}\right) \dot{\omega}_{k}}{[C] {C_{p,\ns/}} + \sum_{k=1}^{\ns/  - 1} \left({C_{p, k}} - {C_{p,\ns/}}\right) [C]_{k}} &\text{\dconp/}\\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns/  - 1} \left(U_{k} - \frac{W_{k} U_{\ns/}}{W_{\ns/}}\right) \dot{\omega}_{k}}{[C] {C_{v,\ns/}} + \sum_{k=1}^{\ns/  - 1} \left({C_{v, k}} - {C_{v,\ns/}}\right) [C]_{k}} &\text{\dconv/}
\end{align}
\end{subequations}
This form of the temperature source term will be used for derivation of Jacobian terms, but~\cref{e:temperature_incomplete} will often be used as well due to its compactness.

@@

\subsection{Reaction rate expressions}
\label{s:rop_net}
The net species production rate for species $k$ is defined as:
\begin{equation}
 \label{e:spec_rop}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr//} \nu_{k,i} q_{i}
\end{equation}
where $\nr//$ is the number of reactions in the chemical kinetic model, $\nu_{k, i}$ the overall stoichiometric coefficient of species $k$ in reaction $i$ and $q_i$ the net rate-of-progress for reaction $i$:
\begin{align}
\nu_{k,i} &= \nu^{\prime\prime}_{k,i} - \nu^{\prime}_{k,i} \\
q_{i} &= R_{i} c_{i}
\end{align}
with $\nu^{\prime}_{k,i}$ and $\nu^{\prime\prime}_{k,i}$ the product and reactant stoichiometric cofficients (respectively) of species $k$ in reaction $i$.
The base rate-of-progress for the $i$th reversible reaction $R_{i}$ is given by:
\begin{align}
R_{i} &= {R_{f,i}} - {R_{r,i}} \label{e:ropnet}\\
{R_{f, i}} &= {k_{f, i}} \prod_{k=1}^{\ns/} [C]_{k}^{\nu^{\prime}_{k,i}} \label{e:ropf}\\
{R_{r, i}} &= {k_{r, i}} \prod_{k=1}^{\ns/} [C]_{k}^{\nu^{\prime\prime}_{k,i}} \label{e:ropr}
\end{align}
where ${k_{f, i}}$ and ${k_{r, i}}$ are the forward and reverse reaction coefficients (respectively) for the $i$th reaction, and the third-body\slash pressure modification $c_{i}$ is given by:
\begin{equation}
\label{e:rxn_pressure}
c_i = \begin{dcases}
  \rhs{pmod}{1}[c] &\text{for elementary reactions,} \\
  \rhs{pmod}{[X]_i}[c] &\text{for third-body enhanced reactions,} \\
  \rhs{pmod}{\frac{P_{r,i}}{1 + P_{r,i}} F_i} &\text{for unimolecular\slash~recombination fall-off reactions, and} \\
  \rhs{pmod}{\frac{1}{1 + P_{r,i}} F_i} &\text{for chemically-activated bimolecular reactions,}
  \end{dcases}
\end{equation}
where for the $i$th reaction $[X]_i$ is the third-body concentration, $P_{r,i}$ is the reduced pressure, and $F_i$ is the falloff blending factor.
These terms are defined in the following sections.

The forward reaction rate coefficient $k_{f, i}$ is given by the three-parameter Arrhenius expression:
\begin{equation}
  \label{e:arrhenius}
  {k_{f, i}} = A_{i} T^{\beta_{i}} \operatorname{exp}\left({- \frac{{E_{a}}_{i}}{T \Ru/}}\right) \;,
\end{equation}
where $A_i$ is the pre-exponential factor, $\beta_i$ is the temperature exponent, and $T_{a, i}$ is the activation temperature given by $T_{a, i} = E_{a, i} / \mathcal{R}$.

@@

\subsection{Reverse rate coefficient}
By definition, the reverse rate coefficient, ${k_{r, i}}$, of irreversible reactions is zero, while reversible reactions may have non-zero ${k_{r, i}}$.
Note that in \texttt{pyJac}, reversible reactions with explicit reverse Arrhenius parameters are split into two irreversible reactions; this simplifies calculation inside the generated code, and eases comparison to Cantera~\cite{Goodwin:2015aa} which applies the same transformation.
For reversible reactions without an explicit parameterization, the reverse rate coefficient is calculated from ratio of the forward rate coefficient and the equilibrium constant:
\begin{align}
 {k_{r, i}} &= \frac{{k_{f, i}}}{{K_{c, i}}}\; , \label{e:kr}\\
 {K_{c, i}} &= {K_{p,i}}\left(\frac{P_{atm}}{T \Ru/}\right)^{\sum_{k=1}^{\ns/} \nu_{k,i}} \; ,\text{ and} \label{e:kc}\\
 {K_{p,i}} &= \text{exp}\left(\frac{\Delta S^{\circ}_i}{\Ru/} - \frac{\Delta H^{\circ}_i}{\Ru/ T}\right) = \text{exp}\left(\sum_{k=1}^{\ns/}\nu_{ki}\left(\frac{S^{\circ}_k}{\Ru/} - \frac{H^{\circ}_k}{\Ru/ T}\right)\right) \label{e:kp}
\end{align}
where $P_{atm}$ is the pressure of one standard atmosphere in appropriate units.

By combining~\cref{e:kc,e:kp}, we obtain:
\begin{equation}
 \label{e:kc_in_kp}
 {K_{c, i}} = \left(\left(\frac{P_{atm}}{\Ru/}\right)^{\sum_{k=1}^{\ns/} \nu_{k,i}}\right) \operatorname{exp}\left({\sum_{k=1}^{\ns/} \nu_{k,i} B_{k}}\right)
\end{equation}
where, $B_k$ is:
\begin{align}
 \label{e:Bk}
 B_{k} &= \frac{S^{\circ}_k}{\Ru/} - \frac{H^{\circ}_k}{\Ru/ T} - \ln{T} \nonumber\; , \\
 B_{k} &= T \left(T \left(T \left(\frac{T a_{k,4}}{20} + \frac{a_{k,3}}{12}\right) + \frac{a_{k,2}}{6}\right) + \frac{a_{k,1}}{2}\right) \nonumber \\
       & \quad + \left(a_{k,0} - 1\right) \log{\left (T \right )} - a_{k,0} + a_{k,6} - \frac{a_{k,5}}{T}
\end{align}
from~\cref{e:hk,e:sk}.

@@

For a reaction enhanced (or diminished) by the presence of a third body, the reaction rate is modified by the third-body concentration $[X]_i$ given by
\begin{equation}
[X]_{i} = \sum_{k=1}^{\ns/} \alpha_{k,i} [C]_{k} \;,
\end{equation}
where $\alpha_{k,i}$ is the third-body efficiency of species $k$ in the $i$th reaction.
For a default third-body efficiency of $\alpha_{k,i} = 1$, this may be rearranged to the compact-storage form:
\begin{equation}
 [X]_{i} = [C] + \sum_{k=1}^{\ns/} \left(\alpha_{k,i} - 1\right) [C]_{k} \;,
\end{equation}
where only species with non-default third-body efficiencies must be stored in the model.
Expanding the concentration of the last species gives:
\begin{equation}
\label{e:thd_mix}
 [X]_{i}=[C] \alpha_{\ns/,i} + \sum_{k=1}^{\ns/  - 1} \left(- \alpha_{\ns/,i} + \alpha_{k,i}\right) [C]_{k}\;.
\end{equation}
This form will be denoted as the mixture-based (or \textbf{mix} for short) third-body efficiency.

If all species in the mixture contribute equally as third bodies, then $\alpha_{k,i} = 1$ for all species:
\begin{equation}
\label{e:thd_unity}
 [X]_{i} = [C] = \frac{P}{\Ru/ T} \;.
\end{equation}
This form will be called the \textbf{unity}-based third-body efficiency.

In addition, a single species may act as the third body in which case
\begin{equation}
 [X]_{i} = [C_m]
\end{equation}
where the $m$th species is the third body.
For evaluation of Jacobian entries, the expanded form:
\begin{equation}
\label{e:thd_spec}
 [X]_{i}=\left([C] - \sum_{k=1}^{\ns/  - 1} [C]_{k}\right) \delta_{\ns/ m} + \left(- \delta_{\ns/ m} + 1\right) [C]_{m}
\end{equation}
will be used, where the Kronecker delta $\delta_{\ns/ m}$ is unity if and only if the third body species $m$ is the last species in the model.
\Cref{e:thd_spec} will be referred to as the \textbf{species}-based third-body efficiency.

@@

\subsection{Temperature source term derivatives}
\label{s:energy_derivs}
\subsubsection{Temperature derivative}
\label{s:denergy_dt}
First, we will derive the term:
\begin{align}
 \mathcal{J}_{1,1} &= \frac{\partial}{\partial T} \frac{\text{d} T}{\text{d} t} \nonumber \\
 \intertext{or more simply:}
 \label{e:dTdotdT_initial}
 \mathcal{J}_{1,1} &= \frac{\partial\dot{T}}{\partial{T}}
\end{align}
To do so, we will more closely follow the \conp/ energy equation derivative with respect to temperature as the process is largely identical between \conp/ and \conv/.

The derivative of a species concentration with respect to temperature for both \conp/ and \conv/ is:
\begin{align}
 \frac{\partial [C]_{k}}{\partial T} =
 \begin{dcases}
  \frac{\partial}{\partial T} \left(\frac{n_k}{V}\right) = 0 & \left\{k \ne \ns/\right\} \\
  \frac{\partial}{\partial T} \left([C] - \sum_{k=1}^{\ns/ - 1}{\frac{n_k}{V}}\right) = -\frac{[C]}{T} & \left\{k = \ns/\right\}
 \end{dcases}
\end{align}
as the pressure and volume are either constants (for \conp/ and \conv/ respectively) or a state variable and a function of time only.

@@

To make the result more managable, the denominator of~\cref{e:temperature_complete} is first collapsed back into the form of~\cref{e:temperature_incomplete}, yielding for \conp/:

@@

 \frac{\partial\dot{T}}{\partial{T}} &= - \frac{\left( [C] \left( \frac{{C_p}_{\ns/}}{T} - \frac{\text{d} {C_p} }{\text{d} T }_{\ns/} \right) - \sum_{k=1}^{-1 + \ns/} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns/} + \frac{\text{d} C_{p, k} }{\text{d} T }\right) [C]_{k}\right)}{\left(\sum_{k=1}^{\ns/} [C]_{k} {C_p}_{k}\right)^{2}} \times \nonumber \\
				     & \mathindent[largerindent] \sum_{k=1}^{-1 + \ns/} \left(H_{k} - \frac{W_{k} H_{\ns/}}{W_{\ns/}}\right) \dot{\omega}_{k} - \nonumber \\
				     & \mathindent \frac{\sum_{k=1}^{-1 + \ns/} \left(\left(H_{k} - \frac{W_{k} H_{\ns/}}{W_{\ns/}}\right) \frac{\partial \dot{\omega}_{k} }{\partial T } + \left(\frac{\text{d} H_{k} }{\text{d} T } - \frac{W_{k}}{W_{\ns/}} \frac{\text{d} H }{\text{d} T }_{\ns/}\right) \dot{\omega}_{k}\right)}{\sum_{k=1}^{\ns/} [C]_{k} {C_p}_{k}}

@@

 \frac{\partial\dot{T}}{\partial{T}} &= \frac{1}{\sum_{k=1}^{\ns/} [C]_{k} {C_p}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \times \nonumber \\
				     & \left( [C] \left( \frac{{C_p}_{\ns/}}{T} - \frac{\text{d} {C_p} }{\text{d} T }_{\ns/} \right) - \sum_{k=1}^{-1 + \ns/} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns/} + \frac{\text{d} {C_p}_{k} }{\text{d} T }\right) [C]_{k}\right) - \nonumber \\
				     & \sum_{k=1}^{-1 + \ns/} \left(\left(H_{k} - \frac{W_{k} H_{\ns/}}{W_{\ns/}}\right) \frac{\partial \dot{\omega}_{k} }{\partial T } + \left(\frac{\text{d} H_{k} }{\text{d} T } - \frac{W_{k}}{W_{\ns/}} \frac{\text{d} H }{\text{d} T }_{\ns/}\right) \dot{\omega}_{k}\right) \Biggr]

@@

 [C] = \sum_{k=1}^{\ns/} [C]_{k}
